---
title: "Prokaryotes"
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(gt)
library(here)

source(here("_common.R"))

samples <- 
  read_csv(here("data", "samples.csv"), col_types = cols(location = col_factor(levels = locations))) |>
  mutate(date = make_datetime(year, month, day)) |> 
  mutate(season = make_season(date))

prokaryotes <- 
  read_csv(here("data", "prokaryotes.csv")) |> 
  left_join(samples, by = join_by(sample)) |> 
  mutate_asv_rarity()
```

## Overview

@fig-season-overview shows how *Verrucomicrobiales* and *Flavobacteriales* reach maximum relative abundance during summer 2022, when fecal contamination event happened repeatedly---see page [Enterococci](enterococci.qmd).

```{r}
#| label: fig-season-overview
#| fig-cap: Seasonal evolution of prokaryotic microbiome composition.

prokaryotes |> 
  filter_analysis() |> 
  mutate(
    order =
      fct_lump_n(order, 11) |>
      fct_na_value_to_level("Other") |>
      fct_reorder(reads, .fun = sum, .desc = T) |>
      fct_relevel("Other", after = Inf)
    ) |> 
  ggplot(aes(x = season, fill = order)) +
  geom_bar(aes(weight = reads), position = "fill", color = "black", linewidth = .3) +
  scale_fill_brewer(palette = "Set3", name = NULL) +
  scale_x_discrete(labels = labels_season, name = NULL) +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Relative abundance") +
  theme_classic()
```

```{r}
#| label: fig-location-overview
#| fig-cap: Prokaryotic microbiome composition for each monitored location.

prokaryotes |> 
  filter_analysis() |> 
  mutate(
    order =
      fct_lump_n(order, 11) |>
      fct_na_value_to_level("Other") |>
      fct_reorder(reads, .fun = sum, .desc = T) |>
      fct_relevel("Other", after = Inf)
    ) |> 
  ggplot(aes(x = location, fill = order)) +
  geom_bar(aes(weight = reads), position = "fill", color = "black", linewidth = .3) +
  scale_fill_brewer(palette = "Set3", name = NULL) +
  scale_x_discrete(labels = labels_location, name = NULL) +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Relative abundance") +
  theme_classic()
```

## Rarity

From @fig-rarity we can easily highlight samples with high abundance of rare taxa. CANT95, CANT1, CANT3, CANT11 and CANT12 are potential examples of a community response to contamination and also in establishing new indicator bacteria.

```{r}
#| label: fig-rarity
#| fig-cap: Rarity composition off all samples grouped in consecutive seasons.

prokaryotes |>
  filter_analysis() |>
  ggplot(aes(x = sample, fill = rarity)) +
  geom_bar(aes(weight = reads), position = "fill", color = "black", linewidth = .2) +
  facet_grid(
    . ~ season, scales = "free_x", space = "free_x",
    labeller = as_labeller(c(labels_location, labels_season))
  ) +
  scale_x_discrete(name = NULL) +
  labs(y = "Relative abundance", fill = "Rarity") +
  guides(fill = guide_legend(nrow = 1)) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = .5), legend.position = "bottom")
```
```{r}
prokaryotes |> 
  filter(sample %in% c("CANT95", "CANT1", "CANT3", "CANT11", "CANT12")) |> 
  distinct(sample, location, date, season) |> 
  gt()
```
## Pathogens and fecal bacteria

Sequenced bacteria which were isolated form human feces, bodily fluids, and feces from land mammals.

*Alistipes indistinctus* @nagai2010
*Alistipes onderdonkii* @sakamoto2020
*Alistipes timonensis* @lagier2012
*Bacteroides dorei* @bakir2006
*Bacteroides vulgatus* (aka *Phocaeicola vulgatus*) @arnold1933
*Bacteroides thetaiotaomicron* @arnold1933 @cato1976
*Bacteroides stercoris* @johnson1986
*Bacteroides plebeius* @kitahara2005
*Bacteroides coprocola* @kitahara2005
*Parabacteroides merdae* @sakamoto2006
*Coprobacter fastidiosus* @shkoporov2013
*Bacteroides eggerthii* @holdeman1974
*Prevotella bivia* @shan1990
*Prevotella disiens* @shan1990
*Prevotella buccalis* @shan1990
*Porphyromonas asaccharolytica* @shan1988
*Prevotella amnii* @lawson2008
*Faecalibacterium sp.* @duncan2002
*Butyricicoccus faecihominis* @takada2016
*Intestinimonas timonensis* @durand2017
*Intestinimonas butyriciproducens* @klaring2013

