---
title: "Sample Metadata"
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| output: false

Sys.setlocale("LC_TIME", "en_US.utf8")

library(tidyverse)
library(osmdata)
library(gt)
library(here)

source(here("_common.R"))

samples <-
  read_csv(here("data", "samples.csv"), col_types = cols(location = col_factor(levels = locations))) |> 
  mutate(date = make_datetime(year, month, day)) |> 
  mutate(season = make_season(date))
```

The microbiome monitoring consists of 47 samples. As shown in @fig-samples-map, these were collected in five different locations across Las Canteras and El Confital. The monitoring period spanned for 2 years, it began in July 2021 and ended in April 2023, with a sampling frequency of around 1 month.

## Locations

```{r}
#| label: fig-samples-map
#| fig-cap: Map of Las Palmas de Gran Canaria showing sampling locations.

coastline <-
  opq(getbb("Las Palmas de Gran Canaria, Spain")) |> 
  add_osm_feature(key = "natural", value = "coastline") |> 
  osmdata_sf()

ggplot() +
  geom_sf(data = coastline$osm_lines, alpha = .5) +
  geom_point(
    data = samples,
    aes(x = longitude, y = latitude)
  ) +
  geom_text(
    data = samples,
    aes(x = longitude, y = latitude, label = location),
    hjust = 1.1,
    vjust = .5
  ) +
  coord_sf(xlim = c(-15.46, -15.40), ylim = c(28.13, 28.17)) +
  theme_bw() +
  theme(axis.title = element_blank())
```
Not all locations were consistently sampled, *El Confital 2* and *Playa Chica* were only sampled once during the entire monitoring period. Since a single sample cannot be representative of any location, these samples will be regarded as outliers in the spatial distribution.

```{r}
samples |>
  ggplot(aes(y = fct_rev(location))) +
  geom_bar() +
  scale_y_discrete(labels = labels_location) +
  labs(x = "Number of samples", y = "") +
  theme_classic()
```
## Dates and seasons

Due to the sporadic nature of contamination events and their possibly limited environmental persistence, understanding the temporal distribution of the samples is crucial for the analysis. Moreover, due to the fact that microbial data will be compared across variables from other databases ---[Nutrients](/nutrients.qmd), [Cells](/cells.qmd), etc.---, some temporal binning has to be made. Seasonal three-months intervals were computed from November 2021 to May 2023. This arbitrary segmentation is intended to balance temporal resolution and data availability.

```{r}
samples |> 
  filter_analysis() |>
  ggplot(aes(x = date, y = fct_rev(location), fill = season)) +
  geom_point(shape = 23, color = "black", size = 3) +
  geom_vline(xintercept = seasons_limits, linetype = "dotted") +
  scale_x_datetime(breaks = seasons_limits, date_labels = "%b %Y") +
  scale_y_discrete(labels = labels_location) +
  scale_fill_discrete(labels = labels_season) +
  guides(fill = guide_legend(nrow = 1)) +
  theme_classic() +
  theme(axis.title = element_blank(), legend.title = element_blank(), legend.position = "bottom")
```
