---
title: "Taxonomical composition"
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(gt)
library(here)

source(here("_common.R"))

samples <- 
  read_csv(here("data", "samples.csv"), col_types = cols(location = col_factor(levels = locations))) |>
  mutate(date = make_datetime(year, month, day)) |> 
  mutate(season = make_season(date))

prokaryotes <- 
  read_csv(here("data", "prokaryotes.csv")) |> 
  left_join(samples, by = join_by(sample)) |> 
  mutate_asv_rate()

eukaryotes <- 
  read_csv(here("data", "eukaryotes.csv")) |> 
  left_join(samples, by = join_by(sample)) |> 
  mutate_asv_rate()
```

## How did the microbiome composition change over time?

```{r}
#| label: fig-prokaryotes-season
#| fig-cap: Seasonal evolution of prokaryotic microbiome composition.

prokaryotes |> 
  mutate(
    order =
      fct_lump_n(order, 11) |>
      fct_na_value_to_level("Other") |>
      fct_reorder(reads, .fun = sum, .desc = T) |>
      fct_relevel("Other", after = Inf)
    ) |> 
  ggplot(aes(x = season, fill = order)) +
  geom_bar(aes(weight = reads), position = "fill", color = "black", linewidth = .3) +
  scale_fill_brewer(palette = "Set3", name = NULL) +
  scale_x_discrete(labels = labels_season, name = NULL) +
  scale_y_continuous(labels = scales::percent, expand = c(0, 0)) +
  labs(y = "Bacteria relative abundance") +
  theme_classic()
```

```{r}
#| label: fig-eukaryotes-season
#| fig-cap: Seasonal evolution of eukaryotic microbiome composition.

eukaryotes |> 
  mutate(
    class =
      fct_lump_n(class, 11) |>
      fct_na_value_to_level("Other") |>
      fct_reorder(reads, .fun = sum, .desc = T) |>
      fct_relevel("Other", after = Inf)
    ) |> 
  ggplot(aes(x = season, fill = class)) +
  geom_bar(aes(weight = reads), position = "fill", color = "black", linewidth = .3) +
  scale_fill_brewer(palette = "Set3", name = NULL) +
  scale_x_discrete(labels = labels_season, name = NULL) +
  scale_y_continuous(labels = scales::percent, expand = c(0, 0)) +
  labs(y = "Eukaryotic relative abundance") +
  theme_classic()
```

## Was there a difference in composition between sampling stations?

```{r}
#| label: fig-prokaryotes-location
#| fig-cap: Prokaryotic microbiome composition for each monitored location.

prokaryotes |> 
  filter_analysis() |> 
  mutate(
    order =
      fct_lump_n(order, 11) |>
      fct_na_value_to_level("Other") |>
      fct_reorder(reads, .fun = sum, .desc = T) |>
      fct_relevel("Other", after = Inf)
    ) |> 
  ggplot(aes(x = location, fill = order)) +
  geom_bar(aes(weight = reads), position = "fill", color = "black", linewidth = .3) +
  scale_fill_brewer(palette = "Set3", name = NULL) +
  scale_x_discrete(labels = labels_location, name = NULL) +
  scale_y_continuous(labels = scales::percent, expand = c(0, 0)) +
  labs(y = "Prokaryotic relative abundance") +
  theme_classic()
```

```{r}
#| label: fig-eukaryotes-location
#| fig-cap: Eukaryotes microbiome composition for each monitored location.

eukaryotes |> 
  filter_analysis() |> 
  mutate(
    class =
      fct_lump_n(class, 11) |>
      fct_na_value_to_level("Other") |>
      fct_reorder(reads, .fun = sum, .desc = T) |>
      fct_relevel("Other", after = Inf)
    ) |> 
  ggplot(aes(x = location, fill = class)) +
  geom_bar(aes(weight = reads), position = "fill", color = "black", linewidth = .3) +
  scale_fill_brewer(palette = "Set3", name = NULL) +
  scale_x_discrete(labels = labels_location, name = NULL) +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Eukaryotic relative abundance") +
  theme_classic()
```

## Which samples were the most unusual?

```{r}
#| label: fig-prokaryotes-unusual
#| fig-cap: Prokaryotic composition based on encounter rate.

prokaryotes |>
  ggplot(aes(x = sample, fill = rate)) +
  geom_bar(aes(weight = reads), position = "fill", color = "black", linewidth = .2) +
  facet_grid(
    . ~ season, scales = "free_x", space = "free_x",
    labeller = as_labeller(c(labels_location, labels_season))
  ) +
  scale_x_discrete(name = NULL) +
  scale_y_continuous(labels = scales::percent, expand = c(0, 0)) +
  labs(y = "Prokaryotic relative abundance", fill = "Rate (%)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = .5))
```

```{r}
#| label: fig-eukaryotes-unusual
#| fig-cap: Eukaryotic composition based on encounter rate.

eukaryotes |>
  ggplot(aes(x = sample, fill = rate)) +
  geom_bar(aes(weight = reads), position = "fill", color = "black", linewidth = .2) +
  facet_grid(
    . ~ season, scales = "free_x", space = "free_x",
    labeller = as_labeller(c(labels_location, labels_season))
  ) +
  scale_x_discrete(name = NULL) +
  scale_y_continuous(labels = scales::percent, expand = c(0, 0)) +
  labs(y = "Eukaryotic relative abundance", fill = "Rate (%)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = .5))
```

## Did we find bacteria from human excretions?

Yes! All bacterial species in @tbl-bacteria were detected in the monitoring and are also documented to come from the human gastrointestinal track or other bodily fluids. The most well-known of which is *Escherichia coli*, although there are many more species of bacteria living inside our bodies.

| Species                            | Reference              |
|------------------------------------|------------------------|
| *Alistipes indistinctus*           | @nagai2010             |
| *Alistipes onderdonkii*            | @sakamoto2020          |
| *Alistipes timonensis*             | @lagier2012            |
| *Bacteroides dorei*                | @bakir2006             |
| *Bacteroides vulgatus*             | @arnold1933            |
| *Bacteroides thetaiotaomicron*     | @arnold1933, @cato1976 |
| *Bacteroides stercoris*            | @johnson1986           |
| *Bacteroides plebeius*             | @kitahara2005          |
| *Bacteroides coprocola*            | @kitahara2005          |
| *Bacteroides eggerthii*            | @holdeman1974          |
| *Butyricicoccus faecihominis*      | @takada2016            |
| *Coprobacter fastidiosus*          | @shkoporov2013         |
| *Escherichia-Shigella coli*        | @devanga2018           |
| *Intestinimonas timonensis*        | @durand2017            |
| *Intestinimonas butyriciproducens* | @klaring2013           |
| *Parabacteroides merdae*           | @sakamoto2006          |
| *Porphyromonas asaccharolytica*    | @shan1988              |
| *Prevotella bivia*                 | @shan1990              |
| *Prevotella disiens*               | @shan1990              |
| *Prevotella buccalis*              | @shan1990              |
| *Prevotella amnii*                 | @lawson2008            |

: Species of present bacteria with a known human origin. {#tbl-bacteria}

### Where were these bacteria detected? When?

```{r}
positive_samples <- 
  prokaryotes |> 
  filter(
    genus == "Alistipes" & species %in% c("indistinctus", "onderdonkii", "timonensis") |
    genus == "Bacteroides" & species %in% c(
      "dorei", "vulgatus", "thetaiotaomicron", "stercoris", "plebeius", "coprocola", "eggerthii"
      ) |
    genus == "Butyricicoccus" & species %in% c("faecihominis") |
    genus == "Coprobacter" & species %in% c("fastidiosus") |
    genus == "Escherichia-Shigella" & species %in% c("coli") |
    genus == "Intestinimonas" & species %in% c("timonensis", "butyriciproducens") |
    genus == "Parabacteroides" & species %in% c("merdae") |
    genus == "Prevotella" & species %in% c("bivia", "disiens", "buccalis", "amnii") |
    genus == "Porphyromonas" & species %in% c("asaccharolytica")
  ) |>
  mutate(key = make_key(year, month, location)) |> 
  pull(sample)

samples |> 
  mutate(hb = if_else(sample %in% c(positive_samples), "Positive", "Negative")) |>
  ggplot(aes(x = date, y = fct_rev(location), fill = hb)) +
  geom_point(shape = 23, color = "black", size = 3) +
  scale_x_datetime(date_labels = "%b %Y") +
  scale_y_discrete(labels = labels_location) +
  scale_fill_discrete(labels = labels_season) +
  guides(fill = guide_legend(nrow = 1)) +
  theme_classic() +
  theme(
    axis.title = element_blank(), legend.title = element_blank(),
    legend.position = "inside", legend.position.inside = c(.83, .9),
    legend.background = element_rect(color = "black")
  )
```

### Are there compositional differences between samples?

```{r}
prokaryotes |>
  mutate(hb = if_else(sample %in% c(positive_samples), "Positive", "Negative")) |>
  mutate(
    order =
      fct_lump_n(order, 11) |>
      fct_na_value_to_level("Other") |>
      fct_reorder(reads, .fun = sum, .desc = T) |>
      fct_relevel("Other", after = Inf)
    ) |> 
  ggplot(aes(x = sample, fill = order)) +
  geom_bar(aes(weight = reads), position = "fill", color = "black", linewidth = .2) +
  facet_grid(
    . ~ hb, scales = "free_x", space = "free_x"
  ) +
  scale_x_discrete(name = NULL) +
  scale_y_continuous(labels = scales::percent, expand = c(0, 0)) +
  scale_fill_brewer(palette = "Set3", name = NULL) +
  labs(y = "Prokaryotic relative abundance", fill = "Rate (%)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = .5))
```

```{r}
eukaryotes |>
  mutate(hb = if_else(sample %in% c(positive_samples), "Positive", "Negative")) |>
  mutate(
    class =
      fct_lump_n(class, 11) |>
      fct_na_value_to_level("Other") |>
      fct_reorder(reads, .fun = sum, .desc = T) |>
      fct_relevel("Other", after = Inf)
    ) |> 
  ggplot(aes(x = sample, fill = class)) +
  geom_bar(aes(weight = reads), position = "fill", color = "black", linewidth = .2) +
  facet_grid(
    . ~ hb, scales = "free_x", space = "free_x"
  ) +
  scale_x_discrete(name = NULL) +
  scale_y_continuous(labels = scales::percent, expand = c(0, 0)) +
  scale_fill_brewer(palette = "Set3", name = NULL) +
  labs(y = "Eukaryotic relative abundance", fill = "Rate (%)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = .5))
```

