---
title: "Bacterial indexes"
---

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(here)

source(here("_common.R"))

samples <- 
  read_csv(here("data", "samples.csv"), col_types = cols(location = col_factor(levels = locations))) |>
  mutate(date = make_datetime(year, month, day)) |> 
  mutate(season = make_season(date))

indexes <-
  read_csv(here("data", "prokaryotes.csv")) |> 
  left_join(samples, by = join_by(sample)) |> 
  group_by(sample, location, date, season) |> 
  summarize(
    alpha           = sum(reads[class    == "Alphaproteobacteria"], na.rm = T),
    gamma           = sum(reads[class    == "Gammaproteobacteria"], na.rm = T),
    alteromonas     = sum(reads[genus    == "Alteromonas"],         na.rm = T),
    sar11           = sum(reads[order    == "SAR11 clade"],         na.rm = T),
    oceanospirillum = sum(reads[order    == "Oceanospirillum"],     na.rm = T),
    bacteroidetes   = sum(reads[phylum   == "Bacteroidota"],        na.rm = T),
    nitrospira      = sum(reads[phylum   == "Nitrospirota"],        na.rm = T),
    burkholderiales = sum(reads[order    == "Burkholderiales"],     na.rm = T),
    bacilli         = sum(reads[class    == "Bacilli"],             na.rm = T),
    clostridia      = sum(reads[class    == "Clostridia"],          na.rm = T)
  ) |> 
  mutate(
    index_as   = (alteromonas) / sar11,
    index_aos  = (alteromonas + oceanospirillum) / sar11,
    index_bngb = (bacteroidetes + nitrospira + gamma) / burkholderiales,
    index_bbca = (bacilli + bacteroidetes + clostridia) / alpha
  ) |> 
  pivot_longer(starts_with("index"), names_to = "index", names_prefix = "index_")
```

| Index                | Formula                                                                                                | Reference      |
|----------------------|:------------------------------------------------------------------------------------------------------:|----------------|
| Alte.:SAR11          | $\frac{\text{Alteromonas}}{\text{SAR11}}$                                                              | [@ferrera2016] |
| (Alte.+Ocea.):SAR11  | $\frac{\text{Alteromonas}+\text{Oceanospirillum}}{\text{SAR11}}$                                       | [@ferrera2016] |
| (Bact.+Nitr.+γ):β    | $\frac{\text{Bacteroidetes}+\text{Nitrospira}+\text{Gammaproteobacteria}}{\text{Betaproteobacteria}}$  | [@garrido2014] |
| (Baci+Bact+Clo):α    |  $\frac{\text{Bacilli}+\text{Bacteroidetes}+\text{Clostridia}}{\text{Alphaproteobacteria}}$            |  [@wu2010]     |
 
## Overview

```{r}
indexes |> 
  filter(value > 0) |> 
  ggplot(aes(x = value, fill = index)) +
  geom_density(color = "black", alpha = .5) +
  labs(x = "", y = "Density", fill = "", color = "") +
  scale_x_log10() +
  scale_fill_discrete(labels = labels_index) +
  theme_classic() +
  theme(legend.position = "inside", legend.position.inside = c(.2, .7))
```

### Distribution

```{r}
indexes |>
  filter(value > 0) |> 
  filter_analysis() |> 
  ggplot(aes(y = fct_rev(location), x = value)) +
  geom_boxplot(outliers = F) +
  facet_wrap(~ index, scales = "free_x", labeller = as_labeller(labels_index)) +
  scale_y_discrete(labels = labels_location) +
  labs(x = "", y = "") +
  theme_classic()
```

```{r}
indexes |>
  filter(value > 0) |> 
  filter_analysis() |> 
  ggplot(aes(x = date, y = value, shape = location, color = location)) +
  geom_jitter() +
  facet_wrap(~ index, scales = "free_y", labeller = as_labeller(labels_index)) +
  scale_color_discrete(labels = labels_location) +
  scale_shape_discrete(labels = labels_location) +
  labs(x = "", y = "", shape = "", color = "") +
  theme_classic() +
  theme(legend.position = "bottom")
```

```{r}
indexes |>
  filter(value > 0) |> 
  filter_analysis() |> 
  ggplot(aes(x = season, y = value)) +
  geom_boxplot(outliers = F) +
  facet_wrap(~ index, scales = "free_y", labeller = as_labeller(labels_index)) +
  scale_x_discrete(labels = labels_season) +
  labs(x = "", y = "", shape = "", color = "") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 20, vjust = .5))
```

